Closing the gap between statistical and scientific workflows for improved forecasts in ecology

### 2025-07-25 revision/response_r1.tex

%emw24Jul: I won't flag all of them, but check for dot dot dot and do a good spellcheck before submitting. 
Thank you for highlighting this as it does capture some of the complexities of how to apply this workflow. To address this, we have worked to clarify the data simulation step (\href{file:forecastflows_r1\#lintarget:firstsim}{line \ref*{lin:firstsim}}, \href{file:forecastflows_r1\#lintarget:repeatsim}{line \ref*{lin:repeatsim}}, \href{file:forecastflows_r1\#lintarget:priorPC}{line \ref*{lin:priorPC}}), and now highlight one should develop specific posterior predictive checks for the question of interest (\href{file:forecastflows_r1\#lintarget:howtoPPC}{line \ref*{lin:howtoPPC}}). As we now state, we believe the best current approach is tailored to the research questions and aims, and thus exactly how conservative posterior predictive checks are may vary in different situations which the researcher should consider.

\begin{mybox}
\emph{Furthermore, I think the authors should include a simple illustration of how these retrodiction checks can flag the need for model changes. For instance, I imagine they could take a fitted model from one of the key cited papers like the LPI trend, and simulate data given spatial/phylogenetic structure, but then refit a model without that and show how the model fails to fit the data and how the analysts would identify (visually, with a new figure) and resolve that by adding model structure/complexity. I am a big proponent of prior/posterior predictive checking and do not disagree with their recommended usage, rather I think more detail is needed for the MS. I think a judicious, simple example here could really help the reader interpret the recommendations and start thinking about how to apply it in their studies. Given the importance of this step in the overall thesis of the MS I believe more detail on how an analyst would do this would be worthwhile.}
\end{mybox}

% here I hope we will have a PPC figure
Thank you for this suggestion. We have added a figure in the supplementary material to illustrate a retrodictive check with a missing phylogenetic structure. We now mention this figure several times throughout the manuscript (\href{file:forecastflows_r1\#lintarget:suppfig1}{lines \ref*{lin:suppfig1}}, \href{file:forecastflows_r1\#lintarget:suppfig2}{\ref*{lin:suppfig2}}, \href{file:forecastflows_r1\#lintarget:suppfig3}{\ref*{lin:suppfig3}}, \href{file:forecastflows_r1\#lintarget:suppfig4}{\ref*{lin:suppfig4}}).

@@

\begin{mybox}
\emph{L36: These workflow citations are all unpublished (arXiv). I’m fine with that, but consider citing some from other fields that have gone through peer review. The caption of Figure 2 includes more, suggest putting them here too.}
\end{mybox}

Done.

\begin{mybox}
\emph{L103: What does “change the model structure” mean here? Adding constraints and priors? Removing or adding complexity?}
\end{mybox}

%emw24Jul: If you add the retro check I think this would be a good example, you could also reference the workflow again? In fact here and elsewhere perhaps I think we could say in response: Great point, we realized we could also better reference the example workflow we provide as supplemental files, which we now do XX times in the paper. 
Great point, we realized we could also better reference the example workflow we provide as supplemental files, which we now do (\href{file:forecastflows_r1\#lintarget:exampleworkflow1}{lines \ref*{lin:exampleworkflow1}}, \href{file:forecastflows_r1\#lintarget:exampleworkflow2}{\ref*{lin:exampleworkflow2}}). We also illustrate this point in the retrodictive check figure in the supplementary.

@@

\begin{mybox}
\emph{L143: FYI the fisheries literature has used the terms pre-data post-model etc. for decades now and have implemented this simulation approach. E.g., https://journal.iwc.int/index.php/jcrm/article/view/239/17. I don’t think it’s necessary to cite this but worth being aware of. You might want to re-cite Hilborn and Mangel (1997) here as they explicitly call for more simulation testing in ecology.}
\end{mybox}

Done.

\begin{mybox}
\emph{L201: “..workflow, non…” This sentence partially answers my question about L103 above}
\end{mybox}

Great, as mentioned above we have tried to add more pointers to what we mean and how to apply this workflow through the revised manuscript.

@@

\begin{mybox}
\emph{L225: I like this paragraph and it makes a point that I hadn’t seen written out so clearly.}
\end{mybox}

Thank you!

\begin{mybox}
\emph{L245: I feel like section 4.2 is slightly out of place. I’d rather see more text devoted to retrodictive checks than this. Consider dropping it and focusing on just a conclusion here}
\end{mybox}

We have removed this last section and have expanded our conclusion.

@@

% Note: I am an applied statistician with experience collaborating with ecologists and developing
% statistical methodology for large-scale long-term monitoring programs. The bulk of my work
% is more concerned with explanatory than predictive inference and my knowledge of building
% ecological forecasting models for inference on a global scale is limited. I reviewed the paper
% through this lens.

We are glad that the reviewer found our manuscript well-written. 
We thank them for the feedback, which has improved the manuscript.

\begin{mybox}
\emph{\textbf{Major comments}\\
1. Clarity on inferential goals of “disconnected paradigms”\\
 When the paper discusses “harmonizing both trend estimation and forecasting” what is meant by this? The title of the paper suggests the inferential goals from both approaches to be predictive in nature (i.e., “forecasting trends”), but it seems plausible that the simpler models are developed for more of an explanatory purpose (e.g., investigating relationships between covariates and mean response) rather than predicting or forecasting future trends, which would make the questions being asked by each modeling approach fundamentally different (see, Shmueli 2010). The paper could be improved if more clarity was provided on the inferential goals of both types of modeling paradigms it discusses, and perhaps provide some discussion why they need to be (or if they should be) harmonized if they are in pursuit of different types of inference.}
\end{mybox}

Thank you for this suggestion. We added a paragraph in the introduction to discuss the inferential goals of these two approaches (\href{file:forecastflows_r1\#lintarget:inferentialgoal}{line \ref*{lin:inferentialgoal}}).

@@

\begin{mybox}
\emph{2. Can this workflow be applied to non-Bayesian models\\
The emphasis on simulation-based investigation of model components is highlighted as a key feature of
the workflow, so it seems a generative (or at least partially generative) model is assumed. Does this model need to be Bayesian? Explicit connections to posterior predictive checks and more discussion for how both prior- and posterior-predictive assessment could be used in a non-Bayesian context would strengthen this work. (Also note that prior predictive checks should be made explicit too – see minor comments).}
\end{mybox}

We clarified that some terms are specific to Bayesian inference but that this workflow can be applied more broadly (\href{file:forecastflows_r1\#lintarget:priorPC}{line \ref*{lin:priorPC}}, \href{file:forecastflows_r1\#lintarget:PPC}{line \ref*{lin:PPC}}, \href{file:forecastflows_r1\#lintarget:broadcontext}{line \ref*{lin:broadcontext}}).

\begin{mybox}
\emph{3. Discussion of Calibration:\\
Lines 98-104 seem to be referring to a check on estimating “the model” using synthetic data that are generated from the model (either through optimization or McMC), but there is no mention of various computational approaches that may be taken to fit “the model.” How does the chosen computational strategy fit into the workflow if at all?}
\end{mybox}

This is a good point, and we could have been clearer on different frameworks that the workflow can be used with. We have worked to clarify that in several places in this revision (in particular, see Box). %emw24Jul: Add line numbers? I think there are two places we now say this.... %emw24Jul: try to avoid saying 'not in scope' and 'no space' when possible.

@@

\begin{mybox}
\emph{Further, is the proposed workflow to conduct a full-blown simulation calibration study (i.e., simulate-fit-check for capture) many times and across all possible parameter values? Assuming yes, I’d be interested in how something like that would scale for the types of models that are the focus of the paper? Could the authors provide more guidance on how to conduct such a study in this context?}
\end{mybox}

We have clarified the data simulation step. We believe a first step would be to simulate from one parameter set (\href{file:forecastflows_r1\#lintarget:firstsim}{line \ref*{lin:firstsim}}), and then to repeat the simulations with several parameter sets (\href{file:forecastflows_r1\#lintarget:repeatsim}{line \ref*{lin:repeatsim}}, \href{file:forecastflows_r1\#lintarget:priorPC}{line \ref*{lin:priorPC}}).

\begin{mybox}
\emph{4. Gaps in the workflow\\
A big argument in the paper is that the proposed workflow could lessen discrepancies in results between studies of trends (lines 123-128). However, in the workflow there is no mention of transparent communication of study design, which if ignored in analysis, could result in discrepancies in results for studies with similar research questions. Differences in sampling designs must also be considered prior to integrating different data sources (if integration is even possible). The importance of study design/sampling design and its ties to appropriate modeling is not discussed. Is it implicitly assumed that there is alignment between design, question, and model in this workflow? Some mention of the importance of study design would strengthen the work.}
\end{mybox}

@@

This is a great point. To address it we have added an entire paragraph that discusses the methodological aspects, including study/sampling design, that should be more transparently shared along with the use of this workflow (\href{file:forecastflows_r1\#lintarget:studydesignandmore}{line \ref*{lin:studydesignandmore}}).

\begin{mybox}
\emph{Additionally, the choice of prior and posterior predictive checks (PPC) is non-trivial and depends on the context of the study. There is discussion of these types of simulation-based investigations, but no mention of the importance of developing checks that assess key features of the observed data that are tied to the research question. Acknowledging the complexity of this step, and including some ideas for how to do this in the context of the types of problems discussed in the paper could strengthen the paper.}
\end{mybox}

Thank you for this suggestion. We now say that the checks should be developed specifically for the research question (\href{file:forecastflows_r1\#lintarget:howtoPPC}{line \ref*{lin:howtoPPC}}), and now show a simple example of retrodictive check in supplementary material.

@@

\begin{mybox}
\emph{5. Theoretical examples provided, but no data application of the proposed workflow\\
Sections 3.1 and 3.2 and the Box outline theoretical examples of how the workflow could be applied for trend estimation and forecasting in the context of types of data, but there are no concrete data examples. A comment on how the proposed workflow could be applied to real data and how it would scale with complexity of data and research questions would be a good addition to the paper if it is not feasible to include real-data examples with the general discussion.}
\end{mybox}

Good point. We provide a worked example, but we realized upon receiving this feedback, that we could better reference the example workflow here and in other places, which we now do (\href{file:forecastflows_r1\#lintarget:exampleworkflow1}{lines \ref*{lin:exampleworkflow1}}, \href{file:forecastflows_r1\#lintarget:exampleworkflow2}{\ref*{lin:exampleworkflow2}}).

@@

\begin{mybox}
\emph{Line 117: should “hypothesis” be “hypotheses”?}
\end{mybox}

Yes, and this is now corrected.

\begin{mybox}
\emph{Lines 198-201: if the goal is accurate and relatively-precise forecasts, do all of the nuisance parameters need to be identifiable for a model to be useful? That is, does identifiability of all parameters matter for process-based models with a goal of forecasting?}
\end{mybox}

We believe that the strength of process-based models (as compared to machine learning for example, \href{file:forecastflows_r1\#lintarget:MLvsPBM}{line \ref*{lin:MLvsPBM}}) is their interpretability. Their parameters should represent biological processes, and their values are of interest. The inferential goal of process-based models is both predictive and explanatory (\href{file:forecastflows_r1\#lintarget:inferentialgoal}{line \ref*{lin:inferentialgoal}}), and poor identifiability could reduce this (\href{file:forecastflows_r1\#lintarget:blackbox}{line \ref*{lin:blackbox}}).

%\clearpage
%\bibliography{../forecastflows.bib}

### 2025-07-25 revision/forecastflows_r1.tex
last modifications (last, really?) and supp figure

% \title{Closing the gap between statistical and scientific workflows for improved forecasts in ecology} 
\title{Closing the gap between statistical and scientific workflows for improved inference in ecology} %emw24Jul: Hmm, you say in the letter you changed the title but I am not sure you did. 
\date{\today}
\author{Victor Van der Meersch$^{1*}$, James Regetz$^{2}$, T. Jonathan Davies$^{1,3}$ \& EM Wolkovich$^{1}$}
\maketitle
\noindent $^{1}$ Forest and Conservation Sciences, University of British Columbia, Vancouver, BC V6T 1Z4, Canada\\
$^{2}$ National Center for Ecological Analysis and Synthesis, 1021 Anacapa St, Santa Barbara, CA 93101, United States\\
$^{3}$ Botany, University of British Columbia, Vancouver, BC V6T 1Z4, Canada \\
$^{*}$ \url{mailto:victor.vandermeersch@ubc.ca}\\

@@

\begin{abstract}
% Concerns about increasing biodiversity loss and climate change have led to greater demands for useful ecological models and forecasts. Datasets relevant for developing these models and forecasts have also increased in size and complexity, including in their geographical, temporal and phylogenetic dimensions. New research often suggests that models accounting for these complexities can yield more accurate trends and predictions. We argue, however, that the usual workflows for model fitting in ecology make it difficult to evaluate and compare most current models for several reasons. First, the research community is split between two disconnected paradigms: either using data to fit simple, trend-like models with one or few parameters, or developing forecasting models that include complex, mechanistic, and/or black box submodels only indirectly informed by data. Second, in both cases, models tend to be developed not only in isolation of one another, but also without a coherent framework for linking scientific questions and understanding to statistical and modeling decisions, throughout the entire modeling process. We propose a unified, principled workflow for end-to-end empirical model development that bridges the gap between process-based and statistical approaches, integrates sound statistical and scientific practices, and especially relies on data simulation to inform decisions at multiple steps in the process. We argue that this approach, coupled with a shift toward universal training, more open model sharing, and alignment on common datasets, could transform for ecological modeling.
Concerns about increasing biodiversity loss and climate change have led to greater demands for useful ecological models. Datasets relevant for developing these models have also increased in size and complexity, including in their geographical, temporal and phylogenetic dimensions. New research often suggests that models accounting for these complexities can yield more accurate trends and predictions. 
%emw24Jul: Are you sure you want 'paradigm'? I love the word and the sentence but I think we mean more two 'groups' or 'spheres' -- I would switch and then carry the language you use to the next sentence. 
We argue, however, that the usual workflows for model fitting in ecology make it difficult to evaluate and compare current models for several reasons. First, the research community is split between two disconnected paradigms that prevent uniting a natural connection between trend estimation and forecasting. One research sphere focuses on using data to fit simple, trend-like models with few parameters, which is separated from research developing forecasting models, which often include complex, mechanistic sub-models only indirectly informed by data. Second, in both cases, models tend to be developed without a coherent framework for linking scientific questions and understanding to statistical and modeling decisions throughout the modeling process. To address these challenges we propose a workflow that integrates statistical and scientific practices through clear steps, many of which rely on data simulation to inform decisions in the process, and where forecasting a natural output. We show how this approach, coupled with a shift toward universal training, more open model sharing, and alignment on common datasets, could harmonize currently divided efforts at trend estimation and forecasting to better inform sustainable policies.

@@

%emw24Jul: Same comment as above about 'paradigms' but here I think you mean 'approaches' so I just changed it. 
\llabel{inferentialgoal}These two modeling approaches may appear to pursue independent inferential goals---explanation vs.\ prediction---but in reality this distinction is blurred. Trend estimation may be nominally focused on explanatory outcomes, yet it rarely links clearly to theories that could provide causal inference, usually because it tends to be carried out using simpler models. Moreover,  estimated trends are frequently treated---implicitly and explicitly---as quantities that project forward into the future, taking on a forecasting role. %emw24Jul: Nice!
In contrast, approaches using complex forecasting models are certainly built for predictive purposes, yet these also have an additional agenda. Developing and using mechanistic, process-based models is often justified by the assumption that ``we must understand to predict," and such models are often judged by how faithfully they mimic the real-world processes they represent, which conflates explanation with prediction \citep{Shmueli2010}.

%emw24Jul: I edited something out that I do and you should avoid (or try to edit out when I add it): adding too many 'subjunctive/conditional' phrases. We ARGUE THAT MANY. ... IN LARGE PART mean ... try to pick one (of two here) conditional phrase and drop the rest. 
We argue that many current debates and diverging forecasts are driven by the incoherent and disconnected workflows used today in ecology \citep{Loreau2022, Talis2023, Johnson2024}. Research estimating biodiversity trends has become focused on methodological aspects because the current workflow fails to examine the gap between ideal and available data, and rarely tests for predictive accuracy that could scale up to allow forecasting. At the same time, process-based models developed for forecasting often evolve through the addition of new separate layers or components because the current workflow rarely examines the model as a functioning whole and thus ignore major problems (e.g., non-identifiability, discussed below). Thus it encourages new parts that are often disconnected from the original research aim, its data stream, and the previous scientific insights.

@@

%emw24Jul: Nice stuff here!
This reality should drive researchers to use more robust and coherent methods. But the current workflows combined with the challenges ecologists are facing---both in terms of data complexity and societal needs---instead may lead to persistent problems.
Trend estimation has focused mostly on fitting a model to empirical data (\llabel{quad1}i.e. post-model/post-data, figure \ref{fig:modeldata}b)---without the checks (\llabel{quad2}post-model/pre-data) and likely feedbacks that often highlight uncertainty and related limitations in the model and/or data (figure \ref{fig:modeldata}b). For forecasting, researchers have focused on making predictions with increasingly complex mechanistic models (figure \ref{fig:modeldata}c), frequently obscuring the steps underlying model building and parameterization (\llabel{quad3}and without a clear distinction between pre- and post-data). Researchers often calibrate the different parts of these models separately, and fix some parameter values based on experiments and expert knowledge, to avoid problems when trying to fit the model as a whole.
Addressing these problems while accounting for the realities of working with ecological data requires a more comprehensive workflow.

We argue a workflow that moves \llabel{quad4}along the data-model space in a coherent sequence of steps (figure \ref{fig:modeldata}d) could reduce many of these problems and thus improve ecological science.
The first step of this workflow is to define an explicit research question and formulate hypotheses (step 1, figure \ref{fig:workflow}). This involves making clear assumptions about the most influential drivers, within the specific context of our study. This should guide the construction of a narrative model of how we believe the system works, focusing on the mechanisms that could generate the data we observe, including the observational error (\llabel{quad5}pre-model/pre-data, figure \ref{fig:modeldata}d).  
\llabel{causalinf}This step of carefully formulating hypotheses has gained increasing attention in the ecological literature \citep{Grace2020}. From this narrative, we can then develop a mathematical model---an ensemble of equations that encapsulates our knowledge and is designed to answer our research question (step 2, figure \ref{fig:workflow}). Generally, starting with a relatively simple model that we could refine later makes understanding the model and how it interacts with empirical data easier (see example workflow we provide). At this stage, prioritizing biologically meaningful parameters is crucial, as it allows us to have a sense of plausible parameter values. This means choosing a model formulation where each parameter corresponds to an interpretable behavior (which sometimes requires considering alternative parameterizations). 
%emw24Jul: 'see example workflow we provide' and 'see example workflow' -- can we standardize our mention of this and make it clear where it is? Like: `see example workflow we provide as supplement files.'

With a model in place, the next step focuses on testing and understanding it via data simulation (step 3, figure \ref{fig:workflow}). `Fake' or `test' data are generated directly from the model by fixing parameters \llabel{firstsim}to some reasonable values, which is straightforward if the parameters are interpretable, and from `fake' (simulated) predictor data. 
We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values (\llabel{quad6}post-model/pre-data, figure \ref{fig:modeldata}d). At this stage, the focus is on understanding the model, so we may need to spend time thinking about whether each parameter is realistic or not. This is also a step for making sure the model is working as expected (see example worklow). \llabel{repeatsim}Ideally, this data simulation step should be repeated several times with different parameter values, sampled within a reasonable range. \llabel{priorPC}In a Bayesian framework, simulated parameters can be sampled from the prior distribution (allowing us to check the implications of our priors \citealp{Gelman2020}), while in other frameworks they can be sampled from a range of biologically reasonable values (which is very similar to the concept of a prior).

@@

Once we are confident about our model structure, we can introduce real data as part of an initial model fitting step (step 4, figure \ref{fig:workflow}). This way, we obtain parameter estimates constrained by observations (\llabel{quad7}post-model/post-data, figure \ref{fig:modeldata}d). 
These parameter estimates lead to the second data simulation step, this time using our fitted model parameters to generate predictions (step 5, figure \ref{fig:workflow}). This---which we call a retrodictive check (\citealp{betanworkflow}, also called a \llabel{PPC}posterior predictive check in a Bayesian workflow,  \citealp{Gelman2020})---allows model output to be compared to observations. \llabel{howtoPPC}This is a powerful approach, in part because checks should be tailored to assess key features of the data and model output together in regards to our research question, but this also means there is no standard way to do this \citep{Gelman2020}.
It is only once all steps have been completed that we can interpret parameter values with respect to our research question.
The workflow encourages a focus on the full model, where any parameter (such as a trend estimate) must be carefully interpreted alongside others, as all are fundamental components that shape both inference and forecasting.

Within such a workflow, forecasting emerges as a natural outcome. Rather than being a final goal, forecasting only involves jointly modeling new circumstances along with the original data. This workflow could also bridge gaps in certain areas between exploratory analyses and developing research questions. For example, in macroecological studies model building is often informed by patterns in the data. The adoption of this workflow would make the exploratory stage more transparent---as an explicit preliminary step for entering the workflow---and would compel researchers to more clearly develop a research question before extensive model fitting. %emw24Jul: possible to ref Fig 1 here?

A key feature of this workflow is the central role of data simulation, which introduces two feedback loops.
The first feedback arises when we evaluate the model on simulated data.
The failure of the model to recover known parameter values and handle the complexity of the simulated data should prompt reconsidering the model, or even reformulating the research question.
Further, this step might reveal that some parameters are highly non-identifiable (meaning the parameter(s) cannot be uniquely estimated),
flagging the need to change the model structure---before incorporating empirical observations. 
The second feedback loop comes from the retrodictive check. 
Discrepancies here may indicate a missing key driver, and suggest the current model is too simplistic. We can refine the model to integrate the missing process(es)---if we can identify them---and return to the start of the workflow. Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake data, such as phylogenetic structure or observational biases (e.g. unbalanced data). For example, if a researcher realizes their empirical data is geographically biased, this bias should be built into the model and thus into this data simulation step. This iterative evaluation of the model moves beyond a simple reliance on goodness-of-fit metrics. At each iteration, we are able to evaluate the model behavior, both with simulated and real data, taking into account our expert knowledge of the ecological processes. \llabel{broadcontext}These feedback loops are applied currently in only in a Bayesian frameworks (to our knowledge), but could and should extend across other modeling approaches given their power to improve models and, in turn, scientific understanding.

@@

%emw24Jul: I removed case study from this draft, I think they're more than case studies and the term no longer helps the paper.
Across the different fields of ecology---for both parameter estimation and forecasting---a systematic application of a coherent workflow could highlight the best opportunities to accurately \llabel{quant1}understand and manage uncertainties and yield new scientific insights. This will help refocus the debate on designing new hypotheses, formulating new questions---and guiding efforts to collect new data. Here, we illustrate how such a workflow could lead to significant improvements in two related areas: (i) estimating global biodiversity trends and (ii) forecasting species and ecosystem dynamics using process-based models.

@@

%emw24Jul: Cite supp retro check figure here below? 
We argue that an improved workflow that required data simulation and retrodictive checks would lead to larger model advances and a greater recognition of uncertainty---thus highlighting likely consistency in estimates across models---that could better aid policy.  Using the workflow would make it more obvious that what now appear as major discrepancies are shifts in point estimates that are generally all in the same uncertainty space \citep{Johnson2024}---and it would challenge modelers to show major predictive advances, which is not currently part of the process. Explanatory power in most models of observational data is usually very low \citep{low2014rising,moller2002much} and thus tests of models' predictions rarely expected. But the workflow highlights that predictions from the model---what we call retrodictive checks---are part of the process of science, and critical to testing for what may be missing in a model.

>>>>

% \title{Closing the gap between statistical and scientific workflows for improved forecasts in ecology} 
\title{Closing the gap between statistical and scientific workflows for improved forecasts in ecology} %emw24Jul: Hmm, you say in the letter you changed the title but I am not sure you did. 
\date{\today}
\author{Victor Van der Meersch$^{1*}$, James Regetz$^{2}$, T. Jonathan Davies$^{1,3}$ \& EM Wolkovich$^{1}$}
\maketitle
\noindent $^{1}$ Forest and Conservation Sciences, University of British Columbia, Vancouver, BC V6T 1Z4, Canada\\
$^{2}$ National Center for Ecological Analysis and Synthesis, 1021 Anacapa St, Santa Barbara, CA 93101, United States\\
$^{3}$ Botany, University of British Columbia, Vancouver, BC V6T 1Z4, Canada \\
$^{*}$ \url{mailto:victor.vandermeersch@ubc.ca}\\

@@

\begin{abstract}
% Concerns about increasing biodiversity loss and climate change have led to greater demands for useful ecological models and forecasts. Datasets relevant for developing these models and forecasts have also increased in size and complexity, including in their geographical, temporal and phylogenetic dimensions. New research often suggests that models accounting for these complexities can yield more accurate trends and predictions. We argue, however, that the usual workflows for model fitting in ecology make it difficult to evaluate and compare most current models for several reasons. First, the research community is split between two disconnected paradigms: either using data to fit simple, trend-like models with one or few parameters, or developing forecasting models that include complex, mechanistic, and/or black box submodels only indirectly informed by data. Second, in both cases, models tend to be developed not only in isolation of one another, but also without a coherent framework for linking scientific questions and understanding to statistical and modeling decisions, throughout the entire modeling process. We propose a unified, principled workflow for end-to-end empirical model development that bridges the gap between process-based and statistical approaches, integrates sound statistical and scientific practices, and especially relies on data simulation to inform decisions at multiple steps in the process. We argue that this approach, coupled with a shift toward universal training, more open model sharing, and alignment on common datasets, could transform for ecological modeling.
Concerns about increasing biodiversity loss and climate change have led to greater demands for useful ecological models. Datasets relevant for developing these models have also increased in size and complexity, including in their geographical, temporal and phylogenetic dimensions. New research often suggests that models accounting for these complexities can yield more accurate trends and predictions. 
%emw24Jul: Are you sure you want 'paradigm'? I love the word and the sentence but I think we mean more two 'groups' or 'spheres' -- I would switch and then carry the language you use to the next sentence. 
We argue, however, that the usual workflows for model fitting in ecology make it difficult to evaluate and compare current models for several reasons. First, the research community is split between two disconnected spheres that prevent uniting a natural connection between trend estimation and forecasting. One research sphere focuses on using data to fit simple, trend-like models with few parameters, which is separated from research developing forecasting models, which often include complex, mechanistic sub-models only indirectly informed by data. Second, in both cases, models tend to be developed without a coherent framework for linking scientific questions and understanding to statistical and modeling decisions throughout the modeling process. To address these challenges we propose a workflow that integrates statistical and scientific practices through clear steps, many of which rely on data simulation to inform decisions in the process, and where forecasting a natural output. We show how this approach, coupled with a shift toward universal training, more open model sharing, and alignment on common datasets, could harmonize currently divided efforts at trend estimation and forecasting to better inform sustainable policies.

@@

\llabel{inferentialgoal}These two modeling approaches may appear to pursue independent inferential goals---explanation vs.\ prediction---but in reality this distinction is blurred. Trend estimation may be nominally focused on explanatory outcomes, yet it rarely links clearly to theories that could provide causal inference, usually because it tends to be carried out using simpler models. Moreover,  estimated trends are frequently treated---implicitly and explicitly---as quantities that project forward into the future, taking on a forecasting role. 
In contrast, approaches using complex forecasting models are certainly built for predictive purposes, yet these also have an additional agenda. Developing and using mechanistic, process-based models is often justified by the assumption that ``we must understand to predict", and such models are often judged by how faithfully they mimic the real-world processes they represent, which conflates explanation with prediction \citep{Shmueli2010}.

We argue that many current debates and diverging forecasts are driven by the incoherent and disconnected workflows used today in ecology \citep{Loreau2022, Talis2023, Johnson2024}. Research estimating biodiversity trends has become focused on methodological aspects because the current workflow fails to examine the gap between ideal and available data, and rarely tests for predictive accuracy that could scale up to allow forecasting. At the same time, process-based models developed for forecasting often evolve through the addition of new separate layers or components because the current workflow rarely examines the model as a functioning whole and thus ignore major problems (e.g., non-identifiability, discussed below). Thus, it encourages new parts that are often disconnected from the original research aim, its data stream, and the previous scientific insights.

@@

This reality should drive researchers to use more robust and coherent methods. But the current workflows combined with the challenges ecologists are facing---both in terms of data complexity and societal needs---instead may lead to persistent problems.
Trend estimation has focused mostly on fitting a model to empirical data (\llabel{quad1}i.e. post-model/post-data, figure \ref{fig:modeldata}b)---without the checks (\llabel{quad2}post-model/pre-data) and likely feedbacks that often highlight uncertainty and related limitations in the model and/or data (figure \ref{fig:modeldata}b). For forecasting, researchers have focused on making predictions with increasingly complex mechanistic models (figure \ref{fig:modeldata}c), frequently obscuring the steps underlying model building and parameterization (\llabel{quad3}and without a clear distinction between pre- and post-data). Researchers often calibrate the different parts of these models separately, and fix some parameter values based on experiments and expert knowledge, to avoid problems when trying to fit the model as a whole.
Addressing these problems while accounting for the realities of working with ecological data requires a more comprehensive workflow.

We argue a workflow that moves \llabel{quad4}along the data-model space in a coherent sequence of steps (figure \ref{fig:modeldata}d) could reduce many of these problems and thus improve ecological science.
The first step of this workflow is to define an explicit research question and formulate hypotheses (step 1, figure \ref{fig:workflow}). This involves making clear assumptions about the most influential drivers, within the specific context of our study. This should guide the construction of a narrative model of how we believe the system works, focusing on the mechanisms that could generate the data we observe, including the observational error (\llabel{quad5}pre-model/pre-data, figure \ref{fig:modeldata}d).  
\llabel{causalinf} This step of carefully formulating hypotheses has gained increasing attention in the ecological literature \citep{Grace2020}. From this narrative, we can then develop a mathematical model---an ensemble of equations that encapsulates our knowledge and is designed to answer our research question (step 2, figure \ref{fig:workflow}). Generally, starting with a relatively simple model that we could refine later makes understanding the model and how it interacts with empirical data easier \llabel{exampleworkflow1}(see example workflow we provide as a supplementary file). At this stage, prioritizing biologically meaningful parameters is crucial, as it allows us to have a sense of plausible parameter values. This means choosing a model formulation where each parameter corresponds to an interpretable behavior (which sometimes requires considering alternative parameterization).

With a model in place, the next step focuses on testing and understanding it via data simulation (step 3, figure \ref{fig:workflow}). `Fake' or `test' data are generated directly from the model by fixing parameters \llabel{firstsim}to some reasonable values, which is straightforward if the parameters are interpretable, and from `fake' (simulated) predictor data. 
We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values (\llabel{quad6}post-model/pre-data, figure \ref{fig:modeldata}d). At this stage, the focus is on understanding the model, so we may need to spend time thinking about whether each parameter is realistic or not. This is also a step for making sure the model is working as expected (see example workflow). \llabel{repeatsim} Ideally, this data simulation step should be repeated several times with different parameter values, sampled within a reasonable range. \llabel{priorPC} In a Bayesian framework, simulated parameters can be sampled from the prior distribution (allowing us to check the implications of our priors \citealp{Gelman2020}), while in other frameworks they can be sampled from a range of biologically reasonable values (which is very similar to the concept of a prior).

@@

Once we are confident about our model structure, we can introduce real data as part of an initial model fitting step (step 4, figure \ref{fig:workflow}). This way, we obtain parameter estimates constrained by observations (\llabel{quad7}post-model/post-data, figure \ref{fig:modeldata}d). 
These parameter estimates lead to the second data simulation step, this time using our fitted model parameters to generate predictions (step 5, figure \ref{fig:workflow}). This---which we call a retrodictive check (\citealp{betanworkflow}, also called a \llabel{PPC}posterior predictive check in a Bayesian workflow,  \citealp{Gelman2020})---allows model output to be compared to observations \llabel{suppfig1}(see supplementary figure 1). \llabel{howtoPPC} This is a powerful approach, in part because checks should be tailored to assess key features of the data and model output together in regard to our research question, but this also means there is no standard way to do this \citep{Gelman2020}.
It is only once all steps have been completed that we can interpret parameter values with respect to our research question.
The workflow encourages a focus on the full model, where any parameter (such as a trend estimate) must be carefully interpreted alongside others, as all are fundamental components that shape both inference and forecasting.

Within such a workflow, forecasting emerges as a natural outcome. Rather than being a final goal, forecasting only involves jointly modeling new circumstances along with the original data. This workflow could also bridge gaps in certain areas between exploratory analyses and developing research questions. For example, in macroecological studies model building is often informed by patterns in the data. The adoption of this workflow would make the exploratory stage more transparent---as an explicit preliminary step for entering the workflow---and would compel researchers to more clearly develop a research question before extensive model fitting (figure 1).

A key feature of this workflow is the central role of data simulation, which introduces two feedback loops.
The first feedback arises when we evaluate the model on simulated data.
The failure of the model to recover known parameter values and handle the complexity of the simulated data should prompt reconsidering the model, or even reformulating the research question.
Further, this step might reveal that some parameters are highly non-identifiable (meaning the parameter(s) cannot be uniquely estimated), flagging the need to change the model structure---before incorporating empirical observations. 
The second feedback loop comes from the retrodictive check. 
Discrepancies here may indicate a missing key driver, and suggest the current model is too simplistic. We can refine the model to integrate the missing process(es)---if we can identify them---and return to the start of the workflow. Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake data, such as phylogenetic structure \llabel{suppfig2}(see supplementary figure 1) or observational biases (e.g. unbalanced data). For example, if a researcher realizes their empirical data is geographically biased, this bias should be built into the model and thus into this data simulation step. This iterative evaluation of the model moves beyond a simple reliance on goodness-of-fit metrics. At each iteration, we are able to evaluate the model behavior, both with simulated and real data, taking into account our expert knowledge of the ecological processes. \llabel{broadcontext}These feedback loops are applied currently in only in Bayesian frameworks (to our knowledge), but could and should extend across other modeling approaches given their power to improve models and, in turn, scientific understanding.

@@

Across the different fields of ecology---for both parameter estimation and forecasting---a systematic application of a coherent workflow could highlight the best opportunities to accurately \llabel{quant1}understand and manage uncertainties and yield new scientific insights. This will help refocus the debate on designing new hypotheses, formulating new questions---and guiding efforts to collect new data. Here, we illustrate how such a workflow could lead to significant improvements in two related areas: (i) estimating global biodiversity trends and (ii) forecasting species and ecosystem dynamics using process-based models.

@@

We argue that an improved workflow that required data simulation and retrodictive checks would lead to larger model advances and a greater recognition of uncertainty---thus highlighting likely consistency in estimates across models---that could better aid policy.  Using the workflow would make it more obvious that what now appear as major discrepancies are shifts in point estimates that are generally all in the same uncertainty space \citep{Johnson2024}---and it would challenge modelers to show major predictive advances, which is not currently part of the process. Explanatory power in most models of observational data is usually very low \citep{low2014rising,moller2002much} and thus tests of models' predictions rarely expected. But the workflow highlights that predictions from the model---what we call retrodictive checks---are part of the process of science, and critical to testing for what may be missing in a model \llabel{suppfig3}(see supplementary figure 1).
